---
title: 'Assignment 6: Using Sobol with an ODE'
author: 'Aakriti Poudel'
date: 'February 23, 2026'
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(deSolve)
library(sensitivity)

```

Consider the following model of forest growth (where forest size in measured in units of carbon (C))

-   $dC/dt  = r*C$ for forests where C is below a threshold canopy closure

-   $dC/dt = g*(1- C/K)$ for forests where carbon is at or above the threshold canopy closure

-   and $K$ is a carrying capacity in units of carbon

The size of the forest ($C$), Canopy closure threshold and carrying capacity are all in units of carbon

You could think of the canopy closure threshold as the size of the forest at which growth rates change from exponential to linear You can think of $r$, as early exponential growth rate and $g$ as the linear growth rate once canopy closure has been reached

## [Your task]{style="color:orange"} {.scrollable}

1.  Implement this model in R (as a differential equation)

```{r}
# Source the forest growth differential equation model
source(here::here("assignments/assignment6/dforestgrowth.R"))

dforestgrowth

```

2.  Run the model for 300 years (using the ODE solver) starting with an initial forest size of 10 kg/C, and using the following parameters:

-   canopy closure threshold of 50 kgC
-   $K$ = 250 kg C (carrying capacity)
-   $r$= 0.01 (exponential growth rate before before canopy closure)
-   $g$ = 2 kg/year (linear growth rate after canopy closure)

```{r}
# Set initial conditions
Cinitial <- 10

# Set simulation times (run the model for 300 years)
simtimes <- seq(from = 1, to = 300)

# Set parameters
parms <- list(r = 0.01, g = 2, K = 250, closure = 50)

# Run the ODE solver
result <- ode(y = Cinitial, times = simtimes, func = dforestgrowth, parms = parms)

# Convert to a data frame with named columns
result <- as.data.frame(result)
colnames(result) <- c("time", "C")

```

3.  Graph the results. Here you are graphing the trajectory with the parameters as given (e.g no uncertainty)

```{r}
# Plot forest growth over 300 years
ggplot(result, aes(time, C)) +
  geom_line(color = "chartreuse4", linewidth = 2) +
  labs(x = "Total no. of years", y = "Size of the forest (kgC)",
       title = "Forest growth over 300 Years") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```

4.  Run a sobol global (vary all parameters at the same time) sensitivity analysis that explores how the estimated **maximum forest size** (e.g maximum of $C$ 300 years, varies with these parameters

-   pre canopy closure growth rate ($r$)
-   post-canopy closure growth rate ($g$)
-   canopy closure threshold and carrying capacity($K$)

Assume that parameters are all normally distributed with means as given above and standard deviation of 10% of mean value

```{r}
# Set number of parameter samples
np <- 2000

# Generate first set of parameter samples
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd = 0.2, n = np)
K <- rnorm(mean = 250, sd = 25, n = np)
closure <- rnorm(mean = 50, sd = 5, n = np)
X1 <- cbind.data.frame(r = r, g = g, K = K, closure = closure)

# Generate second set of parameter samples
r <- rnorm(mean = 0.01, sd = 0.001, n = np)
g <- rnorm(mean = 2, sd = 0.2, n = np)
K <- rnorm(mean = 250, sd = 25, n = np)
closure <- rnorm(mean = 50, sd = 5, n = np)
X2 <- cbind.data.frame(r = r, g = g, K = K, closure = closure)

# Fix any negative values (negatiove values are not meaningful)
X1 <- X1 %>% map_df(pmax, 0.0)
X2 <- X2 %>% map_df(pmax, 0.0)

# Create the Sobol object
sens_C <- sobolSalt(model = NULL, X1, X2, nboot = 300)

# Add column names to the parameter sets
colnames(sens_C$X) <- c("r", "g", "K", "closure")

head(sens_C$X)
```

```{r}
# Create a metric function to compute maximum forest size
compute_forest_metrics <- function(result) {
  maxforest <- max(result$C)
  return(list(maxforest = maxforest))
}
```

```{r}
# Create a wrapper function that runs the ODE and extracts metrics
forest_wrapper <- function(r, g, K, closure, Cinitial, simtimes, odefunc, metricfunc) {
  
  # Create parameter list
  parms <- list(r = r, g = g, K = K, closure = closure)
  
  # Run ode solver
  result <- ode(y = Cinitial, times = simtimes, func = odefunc, parms = parms, method = "euler")
  
  # Convert to data frame
  result <- as.data.frame(result)
  colnames(result) <- c("time", "C")
  
  # Compute metrics
  metrics <- metricfunc(result)
  return(metrics)
}

# Test the wrapper with the first parameter set
forest_wrapper(
  r = sens_C$X[1, "r"], g = sens_C$X[1, "g"],
  K = sens_C$X[1, "K"], closure = sens_C$X[1, "closure"],
  Cinitial = Cinitial, simtimes = simtimes,
  odefunc = dforestgrowth, metricfunc = compute_forest_metrics
)
```

```{r}
# Run the wrapper for all Sobol parameter sets using pmap
allresults <- as.data.frame(sens_C$X) %>%
  pmap(forest_wrapper, Cinitial = Cinitial, simtimes = simtimes,
       odefunc = dforestgrowth, metricfunc = compute_forest_metrics)

# Extract results into a data frame
all_result <- allresults %>%
  map_dfr(`[`, c("maxforest"))

```

5.  Graph the results of the sensitivity analysis as a box plot of maximum forest size and record the two Sobol indices (**S** and **T**).

```{r}
# Generate a boxplot of maximum forest size
ggplot(all_result, aes(y = maxforest)) +
  geom_boxplot(fill = "chartreuse4", alpha = 0.6) +
  labs(y = "Maximum size of the forest (kgC)",
       title = "Uncertainty in maximum forest size (Sobol Analysis)") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```

```{r}
# Check the Sobol object about our results
sens_C_maxforest <- sensitivity::tell(sens_C, all_result$maxforest)

# First-order Sobol indices (S)
rownames(sens_C_maxforest$S) <- c("r", "g", "K", "closure")
cat("First-order Sobol indices (S):\n")
print(sens_C_maxforest$S)

# Total Sobol indices (T)
rownames(sens_C_maxforest$T) <- c("r", "g", "K", "closure")
cat("\nTotal Sobol indices (T):\n")
print(sens_C_maxforest$T)

```

```{r}
# Plot metric against the most sensitive parameter
ggplot(all_result, aes(K, maxforest, col = g)) +
  geom_point(size = 0.8) +
  labs(y = "Size of the maximum forest size (kgC)", x = "Carrying capacity K (kgC)",
       title = "Maximum forest size vs parameters") +
  scale_color_viridis_c(option = "inferno") + 
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))

```

6.  In 2-3 sentences, discuss what the results of your simulation might mean. (For example think about how what parameters climate change might influence).

**Answer:** The Sobol sensitivity analysis shows that the carrying capacity has the biggest effect on the uncertainty in the forest’s maximum size. This makes sense because carrying capacity sets the upper limit of how large the forest can grow. Climate change could lower carrying capacity by increasing drought stress and wildfire frequency. It may also affect how quickly trees continue to grow after the forest canopy has fully formed, since changes in rainfall and temperature influence tree growth. These results suggest that management and policy efforts should focus on protecting or increasing the land’s carrying capacity. The better fire management and water conservation could help maintain larger forests and support long term carbon storage.

Submit rendered Quarto pdf with model implementation, graphs and sensitivity analysis and R file with your model

You can work in groups or individually

## Extra Credit

Compute Sobol indices for a second metric: forest size after a 100 years.

```{r}
# Updated metric function that also computes forest size at year 100
compute_forest_metrics_ec <- function(result) {
  maxforest <- max(result$C)
  # Find forest size closest to year 100
  idx100 <- which(result$time == 100)
  if (length(idx100) == 0) {
    idx100 <- which.min(abs(result$time - 100))
  }
  forest100 <- result$C[idx100]
  return(list(maxforest = maxforest, forest100 = forest100))
}

# Updated wrapper using the new metric function
forest_wrapper_ec <- function(r, g, K, closure, Cinitial, simtimes, odefunc, metricfunc) {
  parms <- list(r = r, g = g, K = K, closure = closure)
  result <- ode(y = Cinitial, times = simtimes, func = odefunc, parms = parms, method = "euler")
  result <- as.data.frame(result)
  colnames(result) <- c("time", "C")
  metrics <- metricfunc(result)
  return(metrics)
}

# Run for all Sobol parameter sets
allresults_ec <- as.data.frame(sens_C$X) %>%
  pmap(forest_wrapper_ec, Cinitial = Cinitial, simtimes = simtimes,
       odefunc = dforestgrowth, metricfunc = compute_forest_metrics_ec)

# Extract results
allres_ec <- allresults_ec %>%
  map_dfr(`[`, c("maxforest", "forest100"))

```


```{r}
# Generate a boxplot of forest size at year 100
ggplot(allres_ec, aes(y = forest100)) +
  geom_boxplot(fill = "firebrick4", alpha = 0.6) +
  labs(y = "Forest size at year 100 (kgC)",
       title = "Uncertainty in forest size after 100 years") +
  theme_minimal() +
   theme(plot.title = element_text(face = "bold", size = 14))

```


```{r}
# Sobol indices for forest size at year 100
sens_C_forest100 <- sensitivity::tell(sens_C, allres_ec$forest100)

rownames(sens_C_forest100$S) <- c("r", "g", "K", "closure")
cat("First-order Sobol indices (S) for forest size at year 100:\n")
print(sens_C_forest100$S)

rownames(sens_C_forest100$T) <- c("r", "g", "K", "closure")
cat("\nTotal Sobol indices (T) for forest size at year 100:\n")
print(sens_C_forest100$T)
```

## Grading Rubric
