---
title: "Assignment 4: Almond Profit Model - Informal Sensitivity Analysis"
author: "Emily Miller and Aakriti Poudel"
date: "`r Sys.Date()`"
format: pdf
editor: visual
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(purrr)
library(ggpubr)
library(scales)
```

## Introduction

This analysis examines how uncertainty in almond price and production cost affects annual profit. We use the almond yield anomaly model from Lobell et al. (2006) combined with a simple profit model to assess sensitivity to two key economic parameters.

## Methods

### Model Structure

The profit model calculates net profit using:

-   **Almond yield anomaly** based on February minimum temperature and January precipitation
-   **Price per ton** of almonds (\$/ton)
-   **Production cost** per acre (\$/acre)
-   **Baseline yield** of 1 ton/acre
-   **Farm size** of 1 acre

### Sensitivity Analysis Parameters

We vary two parameters simultaneously:

1.  **Price**: Uniform distribution from \$4,000-6,000 per ton (±20% around \$5,000 baseline)
2.  **Cost**: Normal distribution with mean \$3,000/acre and standard deviation \$500/acre

We generate 200 parameter combinations and run the model for each.

## Load Functions and Data

```{r load}
# Source the model functions
source("Assignment3_YieldFun.R")
source("profit.R")

# Load climate data 
clim_data <- read_delim("../../Data/clim.txt", delim = " ")
```

## Sensitivity Analysis

```{r sensitivity}
# Set number of samples
nsamples <- 200

# Generate parameter samples
set.seed(123)  # for reproducibility

# Price varies uniformly ±20% around baseline
price <- runif(n = nsamples, min = 4000, max = 6000)

# Cost varies normally around mean
cost <- rnorm(mean = 3000, sd = 500, n = nsamples)

# Combine parameters into dataframe
parms <- data.frame(price = price, cost = cost)

# Add parameter ID for tracking
parms$parm_id <- as.character(1:nrow(parms))

# Run model for all parameter combinations using pmap
results <- parms %>% 
  select(price, cost) %>%
  pmap(profit, 
       clim = clim_data,
       actual_yield = 1,
       acres = 1)

# Check first result structure
results[[1]] %>% head()
```

## Extract and Organize Results

```{r extract_results}
# Extract annual profit data with parameter IDs
annual_profit <- map_df(results, `[`, c("year", "net_profit"), .id = "parm_id")

# Join with parameter values
annual_profit <- annual_profit %>%
  left_join(parms, by = "parm_id") %>%
  mutate(year = as.numeric(year),
         net_profit = as.numeric(net_profit))

# Summary statistics
summary_stats <- annual_profit %>%
  group_by(year) %>%
  summarize(
    mean_profit = mean(net_profit),
    median_profit = median(net_profit),
    min_profit = min(net_profit),
    max_profit = max(net_profit)
  )

summary_stats
```

## Results

### Annual Profit Variability Due to Parameter Uncertainty

```{r plot_boxplot, fig.width=8, fig.height=5}
# Boxplot of annual profit by year showing parameter uncertainty
ggplot(annual_profit, aes(x = as.factor(year), y = net_profit, group = year)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Year",
    y = "Net Profit",
    title = "Annual Almond Profit Variability Across Parameter Uncertainty"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Interpretation

This plot helps us understand how profit uncertainty scales with profitability. In 1995, the box spans several hundred million dollars, indicating massive uncertainty when yields are high. In contrast, low-profit years appear as thin lines with minimal variability. This pattern shows that higher yields amplify the effects of economic uncertainty; when climate produces large yields, small differences in price create massive profit swings. In poor yield years, economic parameters matter less because there's simply less product to sell.

\newpage

## Additional Exploration

### Visualize Individual Parameter Sensitivity

```{r plot_sensitivity, fig.width=10, fig.height=4}
# Calculate mean annual profit for each parameter set
mean_annual_profit <- annual_profit %>%
  group_by(parm_id) %>%
  summarize(mean_profit = mean(net_profit)) %>%
  left_join(parms, by = "parm_id")

# Create two plots showing sensitivity to each parameter
p1 <- ggplot(mean_annual_profit, aes(x = price, y = mean_profit, col = cost)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Price ($/ton)",
    y = "Mean Annual Profit",
    title = "Sensitivity to Price"
  ) +
  theme_minimal()

p2 <- ggplot(mean_annual_profit, aes(x = cost, y = mean_profit, col = price)) +
  geom_point(alpha = 0.6) +
  scale_color_viridis_c(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::dollar_format()) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(
    x = "Cost ($/acre)",
    y = "Mean Annual Profit",
    title = "Sensitivity to Cost"
  ) +
  theme_minimal()

# Combine plots
ggarrange(p1, p2, ncol = 2)

```

-   *Price drives profit:* The left plot shows a tight linear relationship between price and mean annual profit. As price increases from \$4,000 to \$6,000 per ton, profit increases proportionally with minimal scatter

-   *Cost Matters Less:* The right plot shows that production cost (x-axis) barely affects profit. The vertical position of points is determined entirely by price (color) - yellow points (high price) cluster at the top, blue-purple points (low price) at the bottom. At any given cost level, price variation creates the full range of profit outcomes.

**Management Implications** Price uncertainty dominates profitability. In high-yield years, price risk becomes critical - a 20% price difference can mean hundreds of millions in profit variation. Cost management, while important, has minimal impact compared to price exposure. Producers should prioritize price hedging strategies over cost reduction efforts.
